Daily goals

### 2020-06-24 2-body.Rmd
methods specifications

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
- The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the remaining sequences on the alignment.
  The BLAST command line tools [@camacho2009blast] are used for both web-based and local-database searches.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- A pairwise alignment-against-all BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accesion numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the physcraper folder, allowing for secondary analyses to run significantly faster.
- The DNA sequence similarity search can be done on a local database that is easily
  setup by the user. In this case, the BLASTn algorithm is used to performs the similarity search.
- The search can also be performed remotely, on the NCBI database. In this case, the
  bioPython BLAST algorithm is used to perform the similarity search.
- Matched sequences below an e-value, percentage similarity, and outside a minimum
  and maximum length threshold are discarded. This filtering leaves out genomic sequences.
  All acepted sequences are asigned an internal identifier, and are further filtered.
- Because the original alignments usually lack database id numbers, a filtering
  step is needed. Accepted sequences that belong to the
  same taxon of the query sequence, and that are either identical or shorter than
  the original sequence are discarded. Only longer sequences belonging to the
  same taxon as the orignal sequence will be considered further for analysis.
- Among the remaining filtered sequences, there are usually several exemplars per taxon.
  Although it can be useful to keep some of them to, for example, investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
- Reverse complement sequences are identified and translated.
- Users can choose to perform a more "cycles" of sequence similarity search, by blasting the newly found sequences. This can be done iteratively, but by default only sequences in the alignment are blasted. ***Is there an argument to control the number of cycles of blast searches with new sequences?***
- Accepted sequences are downloaded in full, and stored as a local database in a directory that is globally accesible (physcraper/taxonomy), so they are accesible for further runs.
- A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user.

@@

- A gene tree is reconstructed for each alignment provided, using the software RAxML [@stamatakis2014raxml]
  with 100 classic bootstrap [@felsenstein1985confidence] replicates by default. The number of bootsrap replicates can be modified by the user.
  Other type of bootstrap that I think is not yet incorporated into physcraper is the Transfer Bootstrap Expectation (TBE) recently proposed in @lemoine2018renewing.
- The final result is an updated phylogenetic hypothesis for each of the genes provided in the alignment.
- Tips on all trees generated by physcraper are defined by a taxon name space, allowing to perform comparisons and conflict analyses.
- Robinson Foulds metrics
- Describe what a conflict analysis is: Node by node comparison of the resulting clades compared to
- For the conflict analysis to be meaningful, the root of the tree ineeds to be accurately defined.
- Currently, the root is determined by finding the parent node of the sequences that do not belong to the ingroup/ search taxon. This ensures a correct rooting of the tree even when the search taxon is more inclusive than the ingroup.
- Conflict information can only be generated in the context of the whole Open Tree of
  Life. Otherwise, it is not really possible to get conflict data.
***- One way to compare two independent phylogenetic trees is to compare them both to
the synthetic OToL and then measure how well they do against each other***

@@

>>>>

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
- The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the accepted sequences on the alignment.
  The blastn function from the BLAST command line tools [@camacho2009blast] is used for local-database searches.
  A modified biopython blast function is used for web-based searches.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- The DNA sequence similarity search can be done on a local database that is easily
  setup by the user. In this case, the blastn function is used to performs the similarity search [@camacho2009blast].
- The search can also be performed remotely, on the NCBI database. In this case, the
  bioPython BLAST function was modified to accepts is used to perform the similarity search.
- A pairwise alignment-against-all BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accesion numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the physcraper folder, allowing for secondary analyses to run significantly faster.
- Matched sequences below an e-value, percentage similarity, and outside a minimum
  and maximum length threshold are discarded. ***REPORT THE DEFAULT VALUES AND DESCRIBE WHAT THEY MEAN*** This filtering leaves out genomic sequences.
  All acepted sequences are asigned an internal identifier, and are further filtered.
- Because the original alignments usually lack database id numbers, a filtering
  step is needed. Accepted sequences that belong to the
  same taxon of the query sequence, and that are either identical or shorter than
  the original sequence are discarded. Only longer sequences belonging to the
  same taxon as the orignal sequence will be considered further for analysis.
- Among the remaining filtered sequences, there are usually several exemplars per taxon.
  Although it can be useful to keep some of them to, for example, investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
- Reverse complement sequences are identified and translated.
- Users can choose to perform a more "cycles" of sequence similarity search, by blasting the newly found sequences. This can be done iteratively, but by default only sequences in the alignment are blasted. ***Is there an argument to control the number of cycles of blast searches with new sequences?***
- Accepted sequences are downloaded in full, and stored as a local database in a directory that is globally accesible (physcraper/taxonomy), so they are accesible for further runs.
- A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user.

@@

- A gene tree is reconstructed for each alignment provided, using a Maximum Likelihood approach implemented with the software RAxML [@stamatakis2014raxml]
  with 100 classic rapid bootstrap [@felsenstein1985confidence] replicates by default. The number of bootsrap replicates can be modified by the user.
  Other type of bootstrap that I think is not yet incorporated into physcraper is the Transfer Bootstrap Expectation (TBE) recently proposed in @lemoine2018renewing.
- Bootstrap results are summarized with Dendropy ADD CITATION
- The final result is an updated phylogenetic hypothesis for each of the genes provided in the alignment.
- Tips on all trees generated by physcraper are defined by a taxon name space, allowing to perform comparisons and conflict analyses.
- Robinson Foulds weighted and unweighted metrics ARE CALCULATED WITH DENDROPY TOO.
- Describe what a conflict analysis is: Node by node comparison of the resulting clades compared to CITE REDELINGS AND HOLDER [@redelings and holder]
- For the conflict analysis to be meaningful, the root of the tree ineeds to be accurately defined.
- A SUGGESTED DEFAULT ROOTING BASED ON THE OPEN TREE TAXONOMY is implemented for now.  DESCRIBE HOW IT WORKS. SAY THAT IT IS A PROBLEM. Automatic rooting is not that smart yet. The best way right now is for users to define outgroups so trees are better rooted.
- Currently, the root is determined by finding the parent node of the sequences that do not belong to the ingroup/ search taxon. This ensures a correct rooting of the tree even when the search taxon is more inclusive than the ingroup.
- Conflict information can only be generated in the context of the whole Open Tree of
  Life. Otherwise, it is not really possible to get conflict data.
***- One way to compare two independent phylogenetic trees is to compare them both to
the synthetic OToL and then measure how well they do against each other***

@@

# Authors' Contributions

# Data Avilability

### 2020-06-24 _TODO.Rmd
refactoring
methods specifications

- [ ] Finding a tree to update
  - [ ]  Using the curator website. Mention that there is a differnce between focal clade and ingroup. Physcraper does not take into account the focal clade from Open Tree. Because that is arbitrarily assigned by a curator. The focal clade is a search MRCA more than a taxonomic MRCA, physcraper's default is to get the ingroup from Otol, which is not the focal group.
  - [ ] The link between the ingroup and the search taxon.

@@

- [ ] check the all-vs-all blast method, whre does it come from?

@@

- [ ] the root
  - [ ] write something about the need to set the root to be able to do comparisons

- [ ]  In the case that only one taxon from the ingroup is matched in both the tree and alignment, the MRCAT for that single taxon would be determined as ***HOW?***

@@

>>>>

- [ ] Do not cite Felsenstein 1985 for bootstrap, just the call that we use for RAxML:
rax_ex, "-T", "{}".format(self.config.num_threads),
                                 "-m", "GTRCAT",
                                 "-s", alignment,
                                 "-p", "1", "-b", "1", "-#", str(num_reps),
                                 "-n", "BOOT{}".format(self.date)
- [ ] Finding a tree to update
  - [ ]  Using the curator website. Mention that there is a differnce between focal clade and ingroup. Physcraper does not take into account the focal clade from Open Tree. Because that is arbitrarily assigned by a curator. The focal clade is a search MRCA more than a taxonomic MRCA, physcraper's default is to get the ingroup from Otol, which is not the focal group.
  - [ ] The link between the ingroup and the search taxon.

@@

- [ ] check the all-vs-all blast method, where does it come from?

@@

- [ ]  In the case that only one taxon from the ingroup is matched in both the tree and alignment, the MRCAT for that single taxon would be determined as ***HOW?***
  - [ ] automatically, it will return the same taxon, even if its a species.
  - [ ] all tips in original tree are mapped to ott ids
  - [ ] sequences in alignment that are not mapped to ott ids will be kept in the alignment but not blast. Would only happen if somebody decides not to map a tip bc it is no longer the taxon they tought based on the new phylogenetic relationships.

@@

- [ ] determining the root
  - [ ] write something about the need to set the root to be able to do comparisons
  - [ ] say that utis hard to set it up automatically and that solutions are being sought

### 2020-06-24 docs/submission-mee-2020-06-24.tex
methods specifications

@@

@@

@@

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
\item
  The Basic Local Alignment Search Tool, BLAST {[}Altschul \emph{et al.} (\protect\hyperlink{ref-altschul1990basic}{1990}); altschul1997gapped{]} is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the remaining sequences on the alignment.
  The BLAST command line tools (Camacho \emph{et al.} \protect\hyperlink{ref-camacho2009blast}{2009}) are used for both web-based and local-database searches.
\item
  A pairwise alignment-against-all BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accesion numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the physcraper folder, allowing for secondary analyses to run significantly faster.
\item
  The DNA sequence similarity search can be done on a local database that is easily
  setup by the user. In this case, the BLASTn algorithm is used to performs the similarity search.
\item
  The search can also be performed remotely, on the NCBI database. In this case, the
  bioPython BLAST algorithm is used to perform the similarity search.
\item
  Matched sequences below an e-value, percentage similarity, and outside a minimum
  and maximum length threshold are discarded. This filtering leaves out genomic sequences.
  All acepted sequences are asigned an internal identifier, and are further filtered.
\item
  Because the original alignments usually lack database id numbers, a filtering
  step is needed. Accepted sequences that belong to the
  same taxon of the query sequence, and that are either identical or shorter than
  the original sequence are discarded. Only longer sequences belonging to the
  same taxon as the orignal sequence will be considered further for analysis.
\item
  Among the remaining filtered sequences, there are usually several exemplars per taxon.
  Although it can be useful to keep some of them to, for example, investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
\item
  Reverse complement sequences are identified and translated.
\item
  Users can choose to perform a more ``cycles'' of sequence similarity search, by blasting the newly found sequences. This can be done iteratively, but by default only sequences in the alignment are blasted. \textbf{\emph{Is there an argument to control the number of cycles of blast searches with new sequences?}}
\item
  Accepted sequences are downloaded in full, and stored as a local database in a directory that is globally accesible (physcraper/taxonomy), so they are accesible for further runs.
\item
  A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  A gene tree is reconstructed for each alignment provided, using the software RAxML (Stamatakis \protect\hyperlink{ref-stamatakis2014raxml}{2014})
  with 100 classic bootstrap (Felsenstein \protect\hyperlink{ref-felsenstein1985confidence}{1985}) replicates by default. The number of bootsrap replicates can be modified by the user.
  Other type of bootstrap that I think is not yet incorporated into physcraper is the Transfer Bootstrap Expectation (TBE) recently proposed in Lemoine \emph{et al.} (\protect\hyperlink{ref-lemoine2018renewing}{2018}).
\item
  The final result is an updated phylogenetic hypothesis for each of the genes provided in the alignment.
\item
  Tips on all trees generated by physcraper are defined by a taxon name space, allowing to perform comparisons and conflict analyses.
\item
  Robinson Foulds metrics
\item
  Describe what a conflict analysis is: Node by node comparison of the resulting clades compared to
\item
  For the conflict analysis to be meaningful, the root of the tree ineeds to be accurately defined.
\item
  Currently, the root is determined by finding the parent node of the sequences that do not belong to the ingroup/ search taxon. This ensures a correct rooting of the tree even when the search taxon is more inclusive than the ingroup.
\item
  Conflict information can only be generated in the context of the whole Open Tree of
  Life. Otherwise, it is not really possible to get conflict data.
  \textbf{\emph{- One way to compare two independent phylogenetic trees is to compare them both to
  the synthetic OToL and then measure how well they do against each other}}
\end{itemize}

@@

>>>>

Manuscript prepared for submission to Methods in Ecology and Evolution
Article type: Application

@@

\hypertarget{summary}{%
\section{Summary}\label{summary}}

@@

\textbf{Keywords}: phylogeny,

@@

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
\item
  The Basic Local Alignment Search Tool, BLAST {[}Altschul \emph{et al.} (\protect\hyperlink{ref-altschul1990basic}{1990}); altschul1997gapped{]} is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the accepted sequences on the alignment.
  The blastn function from the BLAST command line tools (Camacho \emph{et al.} \protect\hyperlink{ref-camacho2009blast}{2009}) is used for local-database searches.
  A modified biopython blast function is used for web-based searches.
\item
  The DNA sequence similarity search can be done on a local database that is easily
  setup by the user. In this case, the blastn function is used to performs the similarity search (Camacho \emph{et al.} \protect\hyperlink{ref-camacho2009blast}{2009}).
\item
  The search can also be performed remotely, on the NCBI database. In this case, the
  bioPython BLAST function was modified to accepts is used to perform the similarity search.
\item
  A pairwise alignment-against-all BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accesion numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the physcraper folder, allowing for secondary analyses to run significantly faster.
\item
  Matched sequences below an e-value, percentage similarity, and outside a minimum
  and maximum length threshold are discarded. \textbf{\emph{REPORT THE DEFAULT VALUES AND DESCRIBE WHAT THEY MEAN}} This filtering leaves out genomic sequences.
  All acepted sequences are asigned an internal identifier, and are further filtered.
\item
  Because the original alignments usually lack database id numbers, a filtering
  step is needed. Accepted sequences that belong to the
  same taxon of the query sequence, and that are either identical or shorter than
  the original sequence are discarded. Only longer sequences belonging to the
  same taxon as the orignal sequence will be considered further for analysis.
\item
  Among the remaining filtered sequences, there are usually several exemplars per taxon.
  Although it can be useful to keep some of them to, for example, investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
\item
  Reverse complement sequences are identified and translated.
\item
  Users can choose to perform a more ``cycles'' of sequence similarity search, by blasting the newly found sequences. This can be done iteratively, but by default only sequences in the alignment are blasted. \textbf{\emph{Is there an argument to control the number of cycles of blast searches with new sequences?}}
\item
  Accepted sequences are downloaded in full, and stored as a local database in a directory that is globally accesible (physcraper/taxonomy), so they are accesible for further runs.
\item
  A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  A gene tree is reconstructed for each alignment provided, using a Maximum Likelihood approach implemented with the software RAxML (Stamatakis \protect\hyperlink{ref-stamatakis2014raxml}{2014})
  with 100 classic rapid bootstrap (Felsenstein \protect\hyperlink{ref-felsenstein1985confidence}{1985}) replicates by default. The number of bootsrap replicates can be modified by the user.
  Other type of bootstrap that I think is not yet incorporated into physcraper is the Transfer Bootstrap Expectation (TBE) recently proposed in Lemoine \emph{et al.} (\protect\hyperlink{ref-lemoine2018renewing}{2018}).
\item
  Bootstrap results are summarized with Dendropy ADD CITATION
\item
  The final result is an updated phylogenetic hypothesis for each of the genes provided in the alignment.
\item
  Tips on all trees generated by physcraper are defined by a taxon name space, allowing to perform comparisons and conflict analyses.
\item
  Robinson Foulds weighted and unweighted metrics ARE CALCULATED WITH DENDROPY TOO.
\item
  Describe what a conflict analysis is: Node by node comparison of the resulting clades compared to CITE REDELINGS AND HOLDER ({\textbf{???}} and holder)
\item
  For the conflict analysis to be meaningful, the root of the tree ineeds to be accurately defined.
\item
  A SUGGESTED DEFAULT ROOTING BASED ON THE OPEN TREE TAXONOMY is implemented for now. DESCRIBE HOW IT WORKS. SAY THAT IT IS A PROBLEM. Automatic rooting is not that smart yet. The best way right now is for users to define outgroups so trees are better rooted.
\item
  Currently, the root is determined by finding the parent node of the sequences that do not belong to the ingroup/ search taxon. This ensures a correct rooting of the tree even when the search taxon is more inclusive than the ingroup.
\item
  Conflict information can only be generated in the context of the whole Open Tree of
  Life. Otherwise, it is not really possible to get conflict data.
  \textbf{\emph{- One way to compare two independent phylogenetic trees is to compare them both to
  the synthetic OToL and then measure how well they do against each other}}
\end{itemize}

@@

\hypertarget{authors-contributions}{%
\section{Authors' Contributions}\label{authors-contributions}}

\hypertarget{data-avilability}{%
\section{Data Avilability}\label{data-avilability}}

### 2020-06-24 render2-submission-mee.Rmd
methods specifications

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    toc: false
    includes:
      # in_header: preamble.tex
      # after_body: [biorxiv_2019.09.28figcaps.tex]
    # fig_caption: no
geometry: margin = 1in
bibliography: paper.bib
csl: submission-mee/methods-in-ecology-and-evolution.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'm gonna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage{caption}
    - \usepackage{blindtext}
    - \usepackage[left]{lineno}
    - \linenumbers
  # - \usepackage{xr}
    # - \usepackage[utf8]{inputenc}
    # - \usepackage[english]{babel}
    # - \usepackage[document]{ragged2e}
# nocite: |
#   @barker2012going, @barker2015new, @burns2014phylogenetics, # @claramunt2015new, @gibb2015new, @Hedges2015, # @hooper2017chromosomal, @Jetz2012, @price2014niche
---

<!-- ``` {r child = '0-header.Rmd'} -->
<!-- ``` -->
\newpage
``` {r child = '1-abstract.Rmd'}
```
\newpage

``` {r child = '2-body.Rmd'}
```
\newpage
\begin{center}
\textsc{References}
\end{center}

>>>>

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    toc: false
geometry: margin = 1in
bibliography: paper.bib
csl: submission-mee/methods-in-ecology-and-evolution.csl
link-citations: yes
linestretch: 2
header-includes:
    - \usepackage{caption}
    - \usepackage{blindtext}
    - \usepackage[left]{lineno}
    - \linenumbers
params:
  author1: Luna L. Sanchez Reyes
  affiliation: School of Natural Sciences, University of California, Merced
---

`r params$affiliations`
Manuscript prepared for submission to Methods in Ecology and Evolution
Article type: Application

\newpage
# Summary

``` {r child = '1-abstract.Rmd'}
```

**Keywords**: phylogeny,

\newpage

``` {r child = '2-body.Rmd'}
```

\newpage
\begin{center}
\textsc{References}
\end{center}
